---
title: "SoilGrids250 --- Import tiles and combine to a raster stack"
author:
- D G Rossiter
- david.rossiter@wur.nl
date: "`r format(Sys.Date(), '%d-%B-%Y')`"
output:
  word_document:
    toc: yes
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
    theme: spacelab
    number_section: yes
    fig_height: 6
    fig_width: 6
    fig_caption: no
params:
  lrc_long: 78
  lrc_lat: 10
  size: 1
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, purl=FALSE, fig.align = 'center')
knitr::opts_chunk$set(cache.extra = R.version.string, comment="")
```

# Introduction

This script combines SG250 tiles into a raster stack that can be used for (among other things) pattern analysis.

# Directories {#dirs}

Set the base directory where the imported tiles are stored, specific to the local file system. 
This is the same as used in R Markdown file `SoilGrids250_WCS_Import.Rmd`.

```{r set.dirs}
(base.dir <- path.expand("~/ds_reference/DSM2025"))
```

These are the base of destination directories built [below](#dest)

# Packages {#pkgs}

```{r load.pkgs}
library(terra)          # raster data, replaces `raster`
```

# Source files

```{r}
# directory for this AOI
AOI.dir.prefix <- paste0("lat", params$lrc_lat, params$lrc_lat+params$size, 
                         "_lon", params$lrc_long-params$size,
                         params$lrc_long)
(files.dir <- paste0(base.dir, "/", AOI.dir.prefix))
length(files.list <- list.files(files.dir, pattern = "*.tif"))
```

# Raster stack

Make a stack from the set of TIF images.

```{r}
rs <- rast(paste0(files.dir, "/", files.list))
sources(rs)
summary(rs)
```

Plot the first nine layers to check the stack:

```{r}
plot(rs[[1:9]])
```

Save the stack in R format.

```{r save.tile}
print(paste("file name:",
      file.out <- paste0(base.dir, "/", AOI.dir.prefix, "_stack.tif")))
writeRaster(rs, filename = file.out,
            datatype="FLT4S",
            overwrite=TRUE,
            gdal ="TFW=YES")
```

