---
title: "SoilGrids250 --- Import via WCS for DSM comparisons"
author:
- D G Rossiter
- david.rossiter@wur.nl
date: "`r format(Sys.Date(), '%d-%B-%Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: spacelab
    number_section: yes
    fig_height: 6
    fig_width: 6
    fig_caption: no
  pdf_document:
    toc: yes
params:
  lrc_long: -76
  lrc_lat: 42
  size: 1
  voi.n: 4
  quantile.n: 4
  depth.n: 1
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, purl=FALSE, fig.align = 'center')
knitr::opts_chunk$set(cache.extra = R.version.string, comment="")
```

# Introduction

[SoilGrids](https://soilgrids.org/) is a system for global digital soil mapping that makes use of global soil profile information and covariate data to model the spatial distribution of soil properties across the globe. SoilGrids250 is a collections of soil property maps at six standard depths at 250 m grid resolution.

SoilGrids250 filenames, procedures etc. are explained in a [FAQ](https://www.isric.org/explore/soilgrids/faq-soilgrids).
The choice of the [Goode Homolosine projection](https://en.wikipedia.org/wiki/Goode_homolosine_projection) is explained in Moreira de Sousa, L., L. Poggio, and B. Kempen. 2019. Comparison of FOSS4G Supported Equal-Area Projections Using Discrete Distortion Indicatrices. ISPRS International Journal of Geo-Information 8(8): 351. https://doi.org/10.3390/ijgi8080351.

Download of SoilGrids layers is most convenient by Web Coverage Service (WCS), see [this description of procedures from ISRIC](https://www.isric.org/web-coverage-services-wcs).

This script creates a tile for a property and depth slice, over a Area of Interest delimited by geographic coordinates.
Tiles are created in EPSG 4326 (WGS84 long/lat) with SoilGrid's nominal 250 m grid resolution.

This tile can then be compared with other PSM products. 

To use this script:

1. Adjust the [directory structure](#dirs) to your system

Steps 2--3 refer to the YAML headers of the R Markdown source document, or external calls with `knitr::render`. You can ask to be prompted for the parameters with the R command at the console:

```{r eval=FALSE}
rmarkdown::render("SoilGrids250_WCS_import.Rmd", output_format = "html_document", 
                  params = "ask")
```

or you can specify them directly, e.g., 

```{r eval=FALSE}
rmarkdown::render("SoilGrids250_WCS_import.Rmd", output_format = "html_document", 
                    params = list(lrc_long = -76, lrc_lat = 42, 
                                  size = 1, quantile.n = 2, voi.n = 3))
```

See the respective sections of the code for the parameter definitions.

2. [Select a property and quantile](#prop) and [select a depth slice](#depth), using the YAML header or by knitting with parameters.

3. [Select an Area of Interest](#aoi), typically a $1 \times 1^\circ$ tile, using the YAML header or by knitting with parameters.

4. Either compile to HTML or PDF ("knit"), or "Run All" within R Markdown, or call `rmarkdown::render` from the R command line, as explained above.

5. The processed tile will be in the directory structure, in a [subdirectory named for the AOI](#save).

# Directories {#dirs}

Set base directories, specific to the local file system. 

1. `base.dir.import`: This is where downloaded large GeoTIFF are located. Because of their size they may be on a separate file system, e.g., removable or networked drive.

2. `base.dir`: This is where the processed SoilGrids250 maps are stored.


```{r set.dirs}
(base.dir.sg <- path.expand("~/ds_reference/Compare_DSM/SoilGrids250/"))
(base.dir.sg.import <- path.expand("~/tmp/Compare_DSM/SoilGrids250/"))
```

These are the base of destination directories built [below](#dest)

# Packages

```{r load.pkgs}
library(sf)             # spatial data types 
library(terra)          # raster data, replaces `raster`
require(rgdal)          # directly access GDAL functions -- deprecated but still works
require(gdalUtilities)
require(XML)
```

[GDAL](https://gdal.org/) is used for spatial data import/export, coordinate systems etc. Check the GDAL installation, and whether the WCS service is available.


```{r check.GDAL, eval=FALSE, purl=FALSE}
gdal() # version
gdal.drivers <- terra::gdal(drivers = TRUE)
ix <- which(gdal.drivers$name  == "WCS")
gdal.drivers[ix, ]
```

# CRS

We want to use geographic coordinates for the tile. But the ISRIC WCS does not seem to serve this -- or at least, I can not figure out how. So, we must download SoilGrids250 in the native Homolosine CRS. For this we need to know the bounding box in that CRS.

This CRS with pseudo-EPSG code 152160 should be added to to the `epsg` file of the PROJ database ^[for example 
`/Library/Frameworks/PROJ.framework/Versions/6/Resources/proj/epsg`] as a final line, as explained [here](https://www.isric.org/explore/soilgrids/faq-soilgrids#How_can_I_use_the_Homolosine_projection).
But for now do not do this, just specify the projection directly.

```{r crs.igh}
crs.igh <- '+proj=igh +lat_0=0 +lon_0=0 +datum=WGS84 +units=m +no_defs'
```

# Parameters

Parameters for this run:

```{r}
print(paste("lrc_long:", params$lrc_long, "; lrc_lat:", params$lrc_lat, "; size:", params$size))
print(paste("voi.n:", params$voi.n, "; depth.n:", params$depth.n))
print(paste("quantile.n:", params$quantile.n))
```

## Variable and quantile {#voi}

Define the variables for the soil property and layer of interest. See [here](https://www.isric.org/explore/soilgrids/faq-soilgrids#What_do_the_filename_codes_mean) for the naming conventions

`Q0.05` - 5% quantile from the Quantile Random Forest (QRF);
`Q0.5` - median of the distribution from the QRF -- note *not* `Q0.50`;
`Q0.95` - 95% quantile from the QRF;
`mean` - mean of the distribution.

```{r q.list}
quantile.list <- c("Q0.05", "Q0.5", "Q0.95", "mean")
```

Here are the properties predicted by SoilGrids250:

```{r voi.list}
voi.list.sg <- c("clay", "silt", "sand", "phh2o", "cec", "soc", "bdod", "cfvo", "nitrogen", "ocd")
```

## Depth slice {#depth}

Depth slices:
3
```{r d.list}
depth.list <- paste0(c("0-5", "5-15", "15-30", "30-60", "60-100", "100-200"),"cm")
```

Set the property, depth and quantile from the YAML or rendering parameters:

```{r select.prop.d.q}
voi <- voi.list.sg[params$voi.n]
depth <- depth.list[params$depth.n]
quantile.sg <- quantile.list[params$quantile.n]
(voi_layer <- paste(voi, depth, quantile.sg, sep="_")) 
```

## Area of Interest (AOI) {#aoi}

The AOI is a tile using WGS84 geographic coordinates. A $1 \times 1^\circ$ allows comparison with POLARIS, but here other sizes can be specified.

Specify the _lower-right corner_ and _tile size_ from the YAML or rendering parameters:

```{r lrc}
tile.lrc <- c(params$lrc_long, params$lrc_lat) # lower-right corner
size <- params$size                # tile dimensions
```

Compute the four corner and the bounding box. Note because of the projection this is somewhat larger than a $1 \times 1^\circ$  tile.


```{r ulc}
tile.ulc <- c(tile.lrc[1]-size, tile.lrc[2]+size)   # lower-right corner
m <- matrix(c(tile.lrc[1]-size, tile.lrc[2]+size,  #ulc
              tile.lrc[1], tile.lrc[2]+size,  #urc
              tile.lrc[1], tile.lrc[2],       #lrc
              tile.lrc[1]-size, tile.lrc[2]  #lcc
              ),
            nrow=4, byrow = TRUE)
m <- rbind(m, m[1,]) # close the polygon
bb.ll <- st_sfc(st_polygon(list(m)))
st_crs(bb.ll) <- 4326
print(bb.ll)
st_boundary(bb.ll)
```

A prefix for directories, to keep AOI results separate.

```{r dir.prefix}
(AOI.dir.prefix <- paste0("lat", tile.lrc[2], tile.ulc[2],
                         "_lon", tile.ulc[1], tile.lrc[1]))
```

# Destination directories {#dest}

Set destination directories, adding to the base directories the variable of interest, quantile, depth. Make sure the directory exists.

```{r dest.dirs}
(dest.dir.sg.import <-  paste0(base.dir.sg.import, AOI.dir.prefix))
if (!dir.exists(dest.dir.sg.import)) {
   dir.create(dest.dir.sg.import, recursive = TRUE)
}
(dest.dir.sg <-  paste0(base.dir.sg, 
                       AOI.dir.prefix))
if (!dir.exists(dest.dir.sg)) {
  dir.create(dest.dir.sg, recursive = TRUE)
}
```


Convert the long/lat bounding box to the SoilGrids250 projection. We want the extreme values in both X and Y, to ensure we cover the whole tile. If we just use the corners we will cut off some parts at the upper-right an lower-left.

```{r bbox}
(bb.igh <- st_transform(bb.ll, crs.igh))         # reproject the polygon
(bb.igh.coords <-  st_coordinates(bb.igh)[,1:2]) # convert to coordinates, we only need 2D
#  convert to a bounding box, must order these as c(ulx, uly, lrx, lry)
(bb.sg <- as.vector(c(min(bb.igh.coords[,"X"]), 
                     max(bb.igh.coords[,"Y"]), 
                     max(bb.igh.coords[,"X"]), 
                     min(bb.igh.coords[,"Y"]))))
```

# Accessing SoilGrids250 WCS with GDAL

Adapted from instrutions by Luis de Souza (ISRIC), see https://git.wur.nl/isric/soilgrids/soilgrids.notebooks/-/blob/master/markdown/wcs_from_R.md.

```{r}
wcs_request <- "DescribeCoverage" 
wcs_path <- paste0("https://maps.isric.org/mapserv?map=/map/",voi,".map") # Path to the WCS. See maps.isric.org
wcs_service = "SERVICE=WCS"
wcs_version = "VERSION=2.0.1"
(wcs <- paste(wcs_path, wcs_service, wcs_version, wcs_request, sep="&"))
l1 <- newXMLNode("WCS_GDAL")
l1.s <- newXMLNode("ServiceURL", wcs, parent=l1)
l1.l <- newXMLNode("CoverageName", voi_layer, parent=l1)
# Save to local disk
xml.out <- "./sg.xml"
saveXML(l1, file = xml.out)
gdalinfo("./sg.xml")
```

Get the coverage:

```{r}
bb.sg 
crs.igh
(wcs <- paste(wcs_path,wcs_service,wcs_version,sep="&"))
l1 <- newXMLNode("WCS_GDAL")
l1.s <- newXMLNode("ServiceURL", wcs, parent=l1)
l1.l <- newXMLNode("CoverageName", voi_layer, parent=l1)
xml.out = "./sg.xml"
saveXML(l1, file = xml.out)
(file.out <- paste0(dest.dir.sg.import, "/", voi_layer, '.tif'))
gdal_translate(xml.out, file.out,
    tr=c(250,250), 
    projwin = bb.sg, projwin_srs = crs.igh, # corners in this CRS
    co=c("TILED=YES","COMPRESS=DEFLATE","PREDICTOR=2","BIGTIFF=YES")
    )
```

Check the result by reading into R, summarizing, and plotting.

```{r get.sg.igh, fig.width=6, fig.height=6, fig.cap="SoilGrids IGH tile"}
r.sg <- terra::rast(file.out)
print(r.sg)
summary(values(r.sg))
terra::plot(r.sg, col=(sp::bpy.colors(50)))
```

The $0$ values in SoilGrids are in fact un-mapped areas, so convert these to `NA`.

```{r get.sg.igh.na, fig.width=6, fig.height=6, fig.cap="SoilGrids IGH tile, with NA"}
#| label: 0-to-NA
quantile(values(r.sg), seq(0,1,by=0.05))
values(r.sg) <- ifelse(values(r.sg) < 1, NA, values(r.sg))
quantile(values(r.sg), seq(0,1,by=0.05), na.rm = TRUE)
terra::plot(r.sg, col=(sp::bpy.colors(50)))
```


# Resample to WGS84 geographic coordinates

Set the CRS and resample to WGS84 geographic coordinates. Use bilinear interpolation -- notice that the values are now fractional, rather than integers as in the original SoilGrids layer.

```{r transform.crs}
st_bbox(r.sg)
st_crs(r.sg)$proj4string
st_crs(r.sg)$epsg       # not in the EPSG database
r.sg.84 <- terra::project(r.sg, "epsg:4326")
quantile(values(r.sg.84), seq(0,1,by=.1), na.rm = TRUE)
st_crs(r.sg.84)$proj4string
st_crs(r.sg.84)$epsg
```

Crop this to the long/lat bounding box -- recall, the Homolosine bounding box is larger:

```{r show.wgs84.tile, fig.width=6.5, fig.height=6.5, fig.cap="SoilGrids WGS84 tile"}
r.sg.84.crop <- crop(r.sg.84, bb.ll)
quantile(values(r.sg.84.crop), seq(0,1,by=0.05), na.rm = TRUE)
terra::plot(r.sg.84.crop, col=(sp::bpy.colors(50)))
```

# Save tile {#save}

Save this map for further processing, e.g., comparing with other DSM products.

Save the tile. Note that the file name includes the property name and depth slice. Specify the float-4 bit datatype and a GeoTIFF "world" file.  Each tile is about 750 kB.

```{r save.tile}
print(paste("file name:",
      file.out <- paste0(dest.dir.sg, "/", voi_layer, '.tif')))
writeRaster(r.sg.84.crop, file = file.out,
                        overwrite=TRUE, datatype="FLT4S",
                        gdal ="TFW=YES")
```

This can now be imported into analysis scripts.

